---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

<!-- badges: start -->
[![R-CMD-check](https://github.com/hypertidy/wkpool/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/hypertidy/wkpool/actions/workflows/R-CMD-check.yaml)
[![CRAN status](https://www.r-pkg.org/badges/version/wkpool)](https://CRAN.R-project.org/package=wkpool)
<!-- badges: end -->
  
# wkpool

Vertex pool topology for geometry handled by [wk](https://github.com/paleolimbot/wk).

## Design philosophy

Simple features throws away topology — every polygon stores its own coordinates, shared boundaries are duplicated, and expensive operations (GEOS) are needed to rediscover this structure which is then again discarded. 

**wkpool** takes a different approach:

- **Segments are atomic**: Everything is edges (vertex pairs). Polygons, linestrings, meshes — all just segments with different grouping.
- **Vertices live in a pool**: Coordinates stored once, referenced by integer ID (`.vx`).
- **Observe, don't correct**: `establish_topology()` represents the truth of your input. Snapping/fixing is a separate, explicit decision.
- **Topology is discoverable**: `merge_coincident()` finds shared vertices. Shared boundaries, neighbour relations — they emerge from the data.
- **Feature provenance**: Segments track which feature they came from (`.feature`), enabling adjacency and boundary queries.

## Installation

From CRAN use

```r
install.packages("wkpool")
```

For development version use

```r
#install.packages("remotes")
remotes::install_github("hypertidy/wkpool")
```


## Usage

```{r usage}
library(wkpool)
library(wk)

# Any wk-handleable geometry
#x <- wk::as_wkb(<wk-compatible>)
x <- wk::as_wkb(c(
   paste0(
     "MULTIPOLYGON (((0 0, 0 1, 0.75 1, 1 0.8, 0.5 0.7, ",
     "0.8 0.6, 0.69 0, 0 0), (0.2 0.2, 0.5 0.2, ",
     "0.5 0.4, 0.3 0.6, 0.2 0.4, 0.2 0.2)))"),
 "MULTIPOLYGON (((0.69 0, 0.8 0.6, 1.1 0.63, 1.23 0.3, 0.69 0)))"
))

# Establish topology (no coordinate merging yet)
pool <- establish_topology(x)

# What do we have?
topology_report(pool)

# Merge coincident vertices (exact match)
merged <- merge_coincident(pool, tolerance = 0)

# Find shared boundaries
shared <- find_shared_edges(merged)

# Internal boundaries (edges shared by exactly 2 features)
internal <- find_internal_boundaries(merged)

# Neighbour relations
neighbours <- find_neighbours(merged, type = "edge")    # share a boundary
neighbours_v <- find_neighbours(merged, type = "vertex") # share any vertex

# Access the raw structure
pool_vertices(merged)  # data.frame: .vx, x, y
pool_segments(merged)  # data.frame: .vx0, .vx1, .feature
```

## Key decisions

| Decision | Choice | Rationale |
|----------|--------|-----------|
| Pool scope | Per-vector, sidecar attribute | No wk core changes needed, n-geometry ready |
| Snapping | Observe don't correct | Represent truth, fixing is separate |
| Vertex identity | Minted `.vx` integer | Survives subset, lighter than UUID |
| Segment identity | Derived from vertex pair | No ID to track, tuple IS identity |
| Primary vctr | Segments | Geometry is what you subset, vertices follow |
| Foundation | vctrs | Principled combine/subset, tidyverse alignment |

## Next steps

- [x] `wk_handle.wkpool()` — round-trip back to wk geometry
- [x] Path reconstruction — derive linestrings/rings from segment connectivity
- [ ] Apply to production package (trip)
- [ ] Triangulation integration (decido) — indexed triangles from same pool
- [ ] Consider C++ once API is stable


## Code of Conduct
  
Please note that the wkpool project is released with a [Contributor Code of Conduct](https://contributor-covenant.org/version/2/1/CODE_OF_CONDUCT.html). By contributing to this project, you agree to abide by its terms.
